---
title: "Project 1: Chess Tournament"
author: "Khyati Naik"
date: "`r Sys.Date()`"
---

In this project, you’re given a text file with chess tournament results where the information has some structure. Your job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database) with the following information for all of the players:  
Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents  
  
For the first player, the information would be:  
Gary Hua, ON, 6.0, 1794, 1605  
  
1605 was calculated by using the pre-tournament opponents’ ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and dividing by the total number of games played.


## Load packages

```{r, message=FALSE}
library(tidyverse)
```


## Read the data file from github

```{r}
#provide the github data path
dt_path <- "https://raw.githubusercontent.com/Naik-Khyati/607_p1_chess/main/data/tournamentinfo.txt"

# read the txt data file
chess_raw_data <-  read.csv(dt_path, header = FALSE, sep = "|")

head(chess_raw_data,12)
```

## Remove the first 4 rows from the dataframe as those are not required

```{r}
chess_dt <- chess_raw_data %>% slice(-1:-4)
head(chess_dt)
```

## Remove rows with --- as they do not contain any data

```{r}
#delete rows with ----
chess_dt <- chess_dt[!grepl("---", chess_dt[ ,1]), ]
head(chess_dt)
```

## Check if column named V11 contains any information/data

```{r}
unique(chess_dt$V11)
```

## Since column V11 does not contan any data, remove it from the dataframe

```{r}
chess_dt <- chess_dt %>% select(-V11)
head(chess_dt)
```


## Perform further data manipulations

Essentially our data is stored in alternate rows. Hence, i will be saving records in even rows as a data frame and records in odd rows into another data frame.  
  
Also, I will only keep relevant columns in the data frame that i will require int he final data frame.

```{r}
chess_dt_odd <- chess_dt[seq(1, nrow(chess_dt), 2), ]

#  select (V2,V3) %>% rename (name = V2, tot_pts = V3)
chess_dt_even <- chess_dt[seq(2, nrow(chess_dt), 2), ] %>% select (V1,V2) %>% rename (state = V1, pre_rt =V2)
```

## Combine odd and even dataframes

```{r}
chess_dt_comb <- cbind(chess_dt_odd, chess_dt_even)
head(chess_dt_comb)
```

## Data cleaning operations on multiple columns in the dataframe
```{r}
#split the pre rating column by character R:
chess_dt_comb <- separate(chess_dt_comb, pre_rt, c("uscf_id", "rating" ), "R:")

#cleaning the rating column further to keep only digits 
chess_dt_comb$pre_rating <- str_extract(chess_dt_comb$rating, "\\d+\\d")

# Convert pre_rating column to numeric
chess_dt_comb$pre_rating <- as.numeric(chess_dt_comb$pre_rating)

# Convert V3 column to numeric
chess_dt_comb$V3 <- as.numeric(chess_dt_comb$V3)

# Remove white space from column V1
chess_dt_comb$V1 <- trimws(chess_dt_comb$V1)

# Remove white space from column V2
chess_dt_comb$V2 <-  trimws(chess_dt_comb$V2)

# Remove white space from column state
chess_dt_comb$state <-  trimws(chess_dt_comb$state)
```


## To get average opponent rating, first create a dataframe name opp_data in long form using gather function

```{r}
opp_data <- chess_dt_comb %>%
  select(V1,V4:V10) %>%
  gather(key='opp_nbr',value="opp_id",V4:V10) %>% arrange(V1)
glimpse(opp_data)
```

## Perform further data manipulations on opponent id column

```{r}
# keep only numeric values in opp_id column
opp_data$opp_id <- str_extract(opp_data$opp_id,"[0-9]+")
```

## Note that opponent id column have NAs in it

```{r}
opp_data %>%
  select(V1,opp_id) %>% 
  summarise_all(funs(sum(is.na(.))))
```

## Remove rows with na as opponent id

```{r}
opp_data <- opp_data %>% drop_na(opp_id)
```

## Check if there are no NAs in the opp_id column

```{r}
opp_data %>%
  select(V1,opp_id) %>%  
  summarise_all(funs(sum(is.na(.)))) 
```


## Perform further data manipulation on opp_data to fetch average pre rating for the opponents

```{r}
# Left join opp_data with chess_dt_comb to fetch the average pre rating for opponents
opp_data <- opp_data %>%
  left_join(chess_dt_comb, c("opp_id" = "V1")) %>% select (V1, pre_rating) %>%
  group_by (V1) %>%
  summarize(avg_opp_rating=  round(mean(pre_rating),0))

glimpse(opp_data)
```

## Create a chess_dt_final dataframe by left join operation on chess_dt_comb with opp_data. Also rename columns V2 and V3 appropriately.

```{r}
chess_dt_final <- chess_dt_comb %>%
  left_join(opp_data, c("V1" = "V1")) %>% select (V2, state, V3, pre_rating,avg_opp_rating) %>% rename(player_name = V2, tot_points = V3)
head(chess_dt_final,100)
```

## Write the csv into the working directory. This csv can be loaded into database for further analysis.

```{r}
write.csv(chess_dt_final,"chess_tournmnt_clean_data.csv",row.names = FALSE)
```